---
title: "st538 Project 1 Week 3"
author: "Philip Ourso"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(GGally)
library(leaps)
#library(graphics)
library(Rmisc)

if(! (packageVersion("ggplot2") >= "2.0.0")) {
  stop("This version of stat_qqline require ggplot2 version 2.0.0 or greater")
}

StatQQLine <- ggproto("StatQQLine", Stat,
                      compute_group = function(data, scales, distribution = qnorm, dparams = list()) {
                        data <- remove_missing(data, na.rm = TRUE, "sample", name = "stat_qqline")
                        y <- quantile(data$sample, c(0.25, 0.75))
                        x <- do.call(distribution, c(list(p = c(0.25, 0.75)), dparams))
                        slope <- diff(y)/diff(x)
                        int <- y[1L] - slope * x[1L]
                        data.frame(slope = slope, intercept = int)
                      },
                      
                      required_aes = c("sample")
)

stat_qqline <- function(mapping = NULL, data = NULL, geom = "abline",
                        position = "identity", na.rm = FALSE, show.legend = NA,
                        distribution = qnorm, dparams = list(),
                        inherit.aes = TRUE, ...) {
  layer(
    stat = StatQQLine, data = data, mapping = mapping, geom = geom,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, distribution = distribution, dparams = dparams, ...)
  )
}
```

## Project \#1 - Interim Report  
  
  After accounting for gender, educational attainment and typical hours worked last week, is there an effect of state residency on total person's income in the Pacific Northwest states of Washington, Oregon and Idaho?  
  The dataset used included 5-year PUMS data for the states of Washington, Oregon and Idaho, for years 2015-2019. Income and housing cost variables were adjusted with the appropriate adjustment factor and string data were recoded in some cases for convenience.   
  Specific variables include:   
  - PINCP: total person's income  
  - MRGP:   
  - SMOCP: selected monthly owner costs  
  - HINCP: household income  
  - WKHP: usual hours worked per week  
  - SPORDER: person number  
  - RNTP: monthly rent  
  - AGEP: age  
  - SEX: gender  
  - SCHL: educational attainment  
  
### Exploratory data analysis  
  
  Boxplots and density plots of adjusted income by state are difficult to read in the presence of outliers, so the scale of observations is restricted to get a better look at the bulk of the data. Idaho exibits a slightly lower median, while Washington has a slightly higher interquartile range.   
  Pairs plots were created to get a quick snapshot of any potential correlations. Due to the size of the dataset however, a subset of 10,000 observations was considered. Furthermore, categorical variables were considered separately to continuous variables. The log transformation of continuous variables was also considered separately.  
  Of the categorical variables, educational attainment exhibited the most obvious trend, with advanced degree holders exhibiting higher personal income. Gender, class of worker and the aforementioned state also showed variation. Reviewing continuous variables revealed the high likelihood of muticollinearity between two pairs of variables: personal income (PINCP) and household income (HINCP), and mortgage payment (MRGP) and selected monthly owner costs (SMOCP), not surprising considering the presumed relationship between those pairs. Household income and mortgage payment were both dropped from consideration.
  Log transformations of continuous variable revealed no obvious signals.  
```{r, echo=F, message=F, warning=F, inlcude=F}
# read data
df = read.csv(unz("ACSPUMS5Y2019_2022-04-16T191614.zip",
                  "ACSPUMS5Y2019_2022-04-16T165733.csv"
                  ), header = T)
dim(df)
head(df,20)

# char's adding col to end?
df = subset(df, select=-c(X))
colnames(df)

# check for missing
df[is.na(df),]

# recode state, gender as string
df$state = ifelse(df$ST == 16, "ID",
                  ifelse(df$ST == 41, "OR",
                         "WA"
                  )
                  )
df$gender = ifelse(df$SEX==1, "Male", "Female")

# adjust for inflation
df = df %>% 
  mutate(pincp_adj = (PINCP*ADJINC)/1000,
         hincp_adj = (HINCP*ADJINC)/1000,
         mrgp_adj = (MRGP*ADJHSG)/1000,
         smocp_adj = (SMOCP*ADJHSG)/1000,
         rntp_adj = (RNTP*ADJHSG)/1000,
         wagp_adj = (WAGP*ADJINC)/1000
         )

# convert school and class of worker to factor
df = df %>%
  mutate(
    schl_fac = factor(SCHL),
    cow_fac = factor(COW),
    rac_fac = factor(RAC1P)
  )

# retain only records in labor force
df = df %>%
  filter(
    ESR != 0,
    ESR != 3,
    ESR != 6,
    WKHP > 0,
    WKHP < 99,
    wagp_adj > 0
  )

# recode education to reduce levels
df = df %>%
  mutate(
    edu = case_when(
      SCHL <= 19 ~ 0, # high school or less
      SCHL == 20 ~ 1, # associates
      SCHL == 21 ~ 2, # bachelors
      SCHL > 21 ~ 3   # master's or doctoral
    ),
    edu = factor(edu)
  )


```
  
```{r, echo=F, message=F, warning=F}

# univariate distributions

# unadjusted box plots
#ggplot(df, aes(x=factor(ST), y=PINCP)) +
#  geom_boxplot(aes(group=ST)) +
#  theme_minimal()

# adjusted income by state
p0 = ggplot(df, aes(x=factor(state), y=wagp_adj)) +
  geom_boxplot(aes(group=state), varwidth=T, outlier.alpha=0.1) +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "State",
    y = "Adjusted wages (thousands of dollars)",
    title = "Adjusted wages by PacNW state"
  )
p0

# view only up to whiskers
ylim = boxplot.stats(df$wagp_adj)$stats[c(1,5)]
p0 + 
  coord_cartesian(ylim=ylim*1.05) +
  labs(
    title = "Adjusted wages by PacNW state (clipped)"
  )

# adjusted income by state
ggplot(df, aes(x=wagp_adj)) +
  geom_density(aes(color=factor(state))
                 ) +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    y = "Density",
    x = "Adjusted wages (thousands of dollars)",
    title = "Adjusted wages by PacNW state"
  )

# adjusted income by gender
ggplot(df, aes(x=gender, y=wagp_adj)) +
  geom_boxplot(aes(group=gender), varwidth=T, outlier.alpha=0.1) +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "State",
    y = "Adjusted wages (thousands of dollars)",
    title = "Adjusted income by gender"
  )

# adjusted income by school
ggplot(df, aes(x=edu, y=wagp_adj)) +
  geom_boxplot(aes(group=edu), varwidth=T, outlier.alpha=0.1) +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Education",
    y = "Adjusted wages (thousands of dollars)",
    title = "Adjusted wages by education"
  )

# adjusted income by work
ggplot(df, aes(x=cow_fac, y=wagp_adj)) +
  geom_boxplot(aes(group=cow_fac), varwidth=T, outlier.alpha=0.1) +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Class of worker",
    y = "Adjusted wages (thousands of dollars)",
    title = "Adjusted wages by class of worker"
  )

# adjusted income by age
ggplot(df, aes(x=AGEP, y=wagp_adj)) +
  geom_hex() +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Age",
    y = "Adjusted wages (thousands of dollars)",
    title = "Adjusted wages by age"
  )

# adjusted income by age
ggplot(df, aes(x=AGEP, y=wagp_adj)) +
  geom_point(alpha=0.1) +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Age",
    y = "Adjusted wages (thousands of dollars)",
    title = "Adjusted wages by age"
  )

# adjusted income by hours/ week
ggplot(df, aes(x=WKHP, y=wagp_adj)) +
  geom_point(alpha=0.1) +
  theme_minimal() + 
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Age",
    y = "Adjusted wages (thousands of dollars)",
    title = "Adjusted wages by age"
  )

# correlation matrix/ pairs

```
  
```{r, echo=F, message=F, warning=F}
# ggpairs categorical vars
ggpairs(subset(df[sample(nrow(df),100000),], 
               select = c(wagp_adj, state, gender, edu, rac_fac ))) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle=90),
    strip.text.y.right = element_text(angle=0)
  ) + 
  labs(
    title = "Adjusted Personal Income vs categorical variables"
  )

# ggpairs continuous vars
ggpairs(subset(df[sample(nrow(df),100000),], 
               select = c(wagp_adj, hincp_adj, AGEP, WKHP, mrgp_adj, smocp_adj, rntp_adj))) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle=90),
    strip.text.y.right = element_text(angle=0)
  ) +
  labs(
    title = "Adjusted Personal Income vs continuous variables"
  )


#pairs(subset(df[sample(nrow(df),1000),], select = -c(SERIALNO)))
```   
  
```{r, echo=F, message=F, warning=F}
# consider log transformations of continuous vars
df.log = subset(df, select=c(wagp_adj, hincp_adj, AGEP, WKHP, mrgp_adj, smocp_adj, rntp_adj))
df.log = df.log %>%
  mutate(wagp_adj.log = log(wagp_adj),
         hincp_adj.log = log(hincp_adj),
         AGEP.log = log(AGEP),
         WKHP.log = log(WKHP),
         mrgp_adj.log = log(mrgp_adj),
         smocp_adj.log = log(smocp_adj),
         rntp_adj.log = log(rntp_adj)
         )

ggpairs(subset(df.log[sample(nrow(df),100000),], 
               select = c(wagp_adj, wagp_adj.log, hincp_adj.log, AGEP.log, 
                          WKHP.log, mrgp_adj.log, smocp_adj.log, rntp_adj.log))) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle=90),
    strip.text.y.right = element_text(angle=0)
  ) +
  labs(
    title = "Adj wages vs cont. variables log transforms"
  )


#pairs(subset(df[sample(nrow(df),1000),], select = -c(SERIALNO)))
```    
  
### Linear models    
  
  Withholding the state variable, for which inference is desired, best subset selection is used to determine which variables to include in the resulting model. Both adjusted R^2 and Cp prefer 37 variables, predominately inflated by the inclusion of most educational attainment (SCHL) and class of worker (COW) indicator variables. Considering BIC reduces this to 34 variables.  
  Taking the BIC-preferred variables as a starting point, a rich model is fit first. A number of variables are subsequently withheld, one at a time, and F-tests performed to compare reduced models to the full model. In all cases we fail to reject the null hypothesis and the full model is retained.    
  
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# best subset selection, leave state out
df.red = subset(df, 
                select = c(wagp_adj, gender, edu, rac_fac, AGEP, WKHP, smocp_adj, rntp_adj) )

regfit.full = regsubsets(wagp_adj ~ ., df.red, nvmax=20)
reg.summary = summary(regfit.full)

tb = tibble(
  p = 1:length(reg.summary$rsq),
  rsq = reg.summary$rsq,
  rss = reg.summary$rss,
  adjr2 = reg.summary$adjr2,
  cp = reg.summary$cp,
  bic = reg.summary$bic,
)

# adjusted R^2
ggplot(tb, aes(x = p,y = adjr2)) +
  geom_line(color = 'cornflowerblue') +
  geom_point(data=tb[tb$p==which.max( reg.summary$adjr2 ),], 
             mapping=aes(x = p,y = adjr2),color='tomato') +
  theme_minimal() +
  labs(
    x = "Number of variables",
    y = "Adjusted R^2",
    title = paste("Best adjusted R^2 occurs at p==", toString(which.max( reg.summary$adjr2 )))
  )

# adjusted Cp
ggplot(tb, aes(x = p,y = cp)) +
  geom_line(color = 'cornflowerblue') +
  geom_point(data=tb[tb$p==which.min( reg.summary$cp ),], 
             mapping=aes(x = p,y = cp),color='tomato') +
  theme_minimal() +
  labs(
    x = "Number of variables",
    y = "Cp",
    title = paste("Best adjusted Cp occurs at p==", toString(which.min( reg.summary$cp )))
  )

# adjusted Cp
ggplot(tb, aes(x = p,y = bic)) +
  geom_line(color = 'cornflowerblue') +
  geom_point(data=tb[tb$p==which.min( reg.summary$bic ),], 
             mapping=aes(x = p,y = bic),color='tomato') +
  theme_minimal() +
  labs(
    x = "Number of variables",
    y = "Cp",
    title = paste("Best adjusted BIC occurs at p==", toString(which.min( reg.summary$bic )))
  )

coef(regfit.full, 12)

```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
lm.full = lm(wagp_adj ~ gender + edu + rac_fac + AGEP + WKHP +  smocp_adj + rntp_adj, df.red)
summary(lm.full)

# consider reduced models
lm.red1 = lm(wagp_adj ~ gender + edu + AGEP + WKHP + smocp_adj + rntp_adj, df.red)
anova(lm.red1, lm.full)

lm.red2 = lm(wagp_adj ~ gender + edu + rac_fac + AGEP + WKHP + smocp_adj, df.red)
anova(lm.red2, lm.full)

lm.red3 = lm(wagp_adj ~ edu + rac_fac + AGEP + WKHP + smocp_adj + rntp_adj, df.red)
anova(lm.red3, lm.full)

lm.red4 = lm(wagp_adj ~ gender + rac_fac + AGEP + WKHP + smocp_adj + rntp_adj, df.red)
anova(lm.red4, lm.full)

lm.full.log = lm(log(wagp_adj) ~ gender + edu + rac_fac + AGEP + WKHP +  smocp_adj + rntp_adj, df.red)
summary(lm.full.log)
```
  
### Model diagnostics    
  
  Proceeding with the full model identified by best subset selection, model diagnostics reveal large residual errors with extreme deviation from normality, indicating a violation of the assumptions of linear regression. The plot of residual error against fitted value shows several concerning issues: a large percentage of the population has residual errors of hundreds of thousands of dollars, predicted wages are negative for a subset of the population and there is a downward trend indicating that the residual error becomes negative as fitted values become larger.   
  Plots of residuals against explanatory variables reveal further concerning trends, as the constant variance assumption appears to be violated, particularly at the lower and upper values of the predictors. Lastly, the QQ-plot exhibits an extreme departure from normality at upper quantiles.  
  
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# residual vs fitted
df.red.fort = fortify(lm.full, df.red)
qplot(.fitted, .resid, data=df.red.fort)

# residual vs explanatory
qplot(AGEP, .resid, data = df.red.fort) 
qplot(WKHP, .resid, data = df.red.fort) 
#qplot(WKWN, .resid, data = df.red.fort) 
qplot(smocp_adj, .resid, data = df.red.fort) 
qplot(rntp_adj, .resid, data = df.red.fort) 
#multiplot(p1, p2, p3, p4, p5, cols=5)

# qqplot
qplot(sample = df.red.fort$.resid) + stat_qqline() 

```
  
### Conclusion  
  
  Clearly the model, as identified by best subset selection and fitted by multiple linear regression, exhibits several serious violations of assumptions. For completeness sake, in order to answer the question posited at the outset, a linear regression model is fit with all previously included explanatory variables and Pacific Northwest state, as a factor. Reviewing the results from this new model indicates that, with respect to Idaho residency as a baseline, only Washington residency is found to be significant at a p-value < 0.0001. The coefficient for the Washington indicator is estimated to be 5.45, indicating a non-trivial impact upon wages (in thousands of dollars).  
  
```{r, echo=False, message=FALSE, warning=FALSE}  
df.red.state = subset(df, 
                select = c(wagp_adj, gender, edu, rac_fac, AGEP, WKHP, smocp_adj, rntp_adj, state) )
lm.full.state = lm(wagp_adj ~ gender + edu + rac_fac + AGEP + WKHP +  smocp_adj + rntp_adj + state, df.red.state)
summary(lm.full.state)  

```  


